The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (9th edition -- distributed with CrypTool version 1.4.20)} \vspace{60pt}

### 2009-07-04 script/en/chapters/numbertheory.tex
chapter on number theory: add Sage examples for calculating primitive roots; change some occurrences of Mathematica, Pari/GP to Sage.

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
we have\footnote{%
  See \hyperlink{AppArith2}{Appendix E of this chapter} for source code implementing the
  square and multiply method in Mathematica and Pari-GP, which can be used
  to reproduce the calculations above. }:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}

@@

{\bf Example 1:} \\
The following table\footnote{%
See \hyperlink{AppArith3a}{Appendix E of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$:

@@

{\bf Example 2:} \\
The following table\footnote{%
See \hyperlink{AppArith3b}{Appendix E of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$:

@@

{\bf Example 4:} \\
The following table\footnote{%
See \hyperlink{AppArith3c}{Appendix E of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} answers the question as to whether the number $7$ is a primitive root of $46$.
The requirement/condition $gcd(7,46)=1$ is fulfilled.

@@

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code to do RSA
encryption using Mathematica and Pari-GP. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix E of this chapter} for the source code to factorise
the number $J(n)$ using Mathematica and Pari-GP.
Using CrypTool\index{CrypTool} you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorised but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix E of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix E of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix E of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt 198967 051405 254571 115318 014251}

@@

\hypertarget{AppArith3a}{}
\subsubsection*{Multiplicative order and primitive roots}

@@

>>>>

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source
code to compute the tables using Sage.}
illustrate the properties of multiplicative inverses.

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
we have\footnote{%
See \hyperlink{AppArith2}{Appendix E of this chapter} for source code
implementing the square and multiply method in Sage, which can be used
to reproduce the calculations above.}:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}

@@

{\bf Example 1:} \\
The following table\footnote{%
See \hyperlink{AppArith3a}{Appendix E of this chapter} for the source
code to generate the table using Sage.} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$:

@@

{\bf Example 2:} \\
The following table\footnote{%
See \hyperlink{AppArith3b}{Appendix E of this chapter} for the source
code to generate the table using Sage.} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$:

@@

{\bf Example 4:} \\
The following table\footnote{%
See \hyperlink{AppArith3c}{Appendix E of this chapter} for the source
code to generate the table using Sage.} answers the question as to
whether the number $7$ is a primitive root of $46$. 
The requirement/condition $gcd(7,46)=1$ is fulfilled.

@@

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code
to do RSA encryption using Sage. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix E of this chapter} for the source
code to factorise the number $J(n)$ using Sage.
Using CrypTool\index{CrypTool} you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorised but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 198967 051405 254571 115318 014251}

@@

\hypertarget{AppArith3a}{}
\subsubsection*{Multiplicative order}

@@

\subsection*{Primitive roots}

Computing a primitive root in Sage is very straightforward. If
\verb!n! is an integer, the command \verb!primitive_root(n)! computes
a primitive root of the multiplicative group
$(\mathbf{Z} / n \mathbf{Z})^{\ast}$, if one exists. Where $n$ is
prime, then this is the same as calculating a primitive root of
$\mathbf{Z} / n \mathbf{Z}$. Here, we calculate some primitive roots
of a few integers.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: primitive_root(4)
3
sage: primitive_root(22)
13
sage: for p in primes(1, 50):
....:     print p, primitive_root(p)
....:     
2 1
3 2
5 2
7 3
11 2
13 2
17 3
19 2
23 5
29 2
31 3
37 2
41 6
43 3
47 5
\end{Verbatim}

If $p$ is prime, then $\mathbf{Z} / p \mathbf{Z}$ has at least one
primitive root. Sometimes we want to compute all the primitive roots
of $\mathbf{Z} / p \mathbf{Z}$, not just any primitive root of
$\mathbf{Z} / p \mathbf{Z}$. The following function can do this.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_primroots(M):
....:     r"""
....:     Return all the primitive roots of M.
....:     """
....:     g = primitive_root(M)
....:     targetOrder = euler_phi(M)
....:     L = []
....:     # Stepping through all odd integers from 1 up to M, not including
....:     # M. So this loop only considers values of i where 1 <= i < M.
....:     for i in xrange(1, M, 2):
....:         testGen = Mod(g^i, M)
....:         if testGen.multiplicative_order() == targetOrder:
....:             L.append(testGen.lift())
....:     # removing duplicates
....:     return Set(L)
\end{Verbatim}

\noindent For example, here is a list of all primitive roots of the prime 541.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: enum_primroots(541)
{2, 517, 10, 523, 13, 14, 527, 528, 18, 531, 24, 539, 30, 37, 40, 51,
54, 55, 59, 62, 65, 67, 68, 72, 73, 77, 83, 86, 87, 91, 94, 96, 98,
99, 107, 113, 114, 116, 117, 126, 127, 128, 131, 132, 138, 150, 152,
153, 156, 158, 163, 176, 181, 183, 184, 195, 197, 199, 206, 208,
210, 213, 218, 220, 223, 224, 244, 248, 250, 257, 258, 259, 260,
261, 263, 267, 269, 270, 271, 272, 274, 278, 280, 281, 282, 283,
284, 291, 293, 297, 317, 318, 321, 323, 328, 331, 333, 335, 342,
344, 346, 357, 358, 360, 365, 378, 383, 385, 388, 389, 391, 403,
409, 410, 413, 414, 415, 424, 425, 427, 428, 434, 442, 443, 445,
447, 450, 454, 455, 458, 464, 468, 469, 473, 474, 476, 479, 482,
486, 487, 490, 501, 504, 511}
\end{Verbatim}

\noindent With a little bit of programming, we can find out how long
it takes to compute all the primitive roots of some number of primes.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def primroots_iter(start, end):
....:     r"""
....:     Compute all primitive roots of all primes between start and end,
....:     inclusive. This uses a primes iterator.
....:     """
....:     for p in primes(start, end+1):
....:         enum_primroots(p)
....:         
sage: %time primroots_iter(1, 1500)
CPU times: user 2.04 s, sys: 0.17 s, total: 2.21 s
Wall time: 2.21 s
\end{Verbatim}

\noindent And we can count how many primes for which to enumerate all
their primitive roots.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def primroots_primetest(start, end):
....:     r"""
....:     Compute all primitive roots of all primes between start and end,
....:     inclusive. This actually perform primality tests, so it allows us
....:     to count the number of primes from start to end, inclusive.
....:     """
....:     nPrimes = 0
....:     for n in xrange(start, end+1):
....:         if is_prime(n):
....:             nPrimes += 1
....:             enum_primroots(n)
....:     print "Enumerate all primitive roots of %s primes." % nPrimes
....:     
sage: %time primroots_primetest(1, 1500)
Enumerate all primitive roots of 239 primes.
CPU times: user 2.35 s, sys: 0.03 s, total: 2.38 s
Wall time: 2.39 s
\end{Verbatim}

Using our custom-defined function \verb!enum_primroots!, we can find
all primitive roots of a prime integer $p$. From this list of primitive
roots, we can determine the smallest and largest primitive root for
$\mathbf{Z} / p \mathbf{Z}$, as well as count the number of primitive
roots of $\mathbf{Z} / p \mathbf{Z}$.
Figure~\ref{fig:primitive_roots_all} graphs the number of primitive
roots for each prime between 1 and 100000. The $x$-axis represents
primes between 1 and 100000, while the $y$-axis counts the number of
primitive roots for each prime within that interval.

\begin{figure}[!htbp]
\centering
\includegraphics{figures/primitive-roots-all}
\caption{The number of primitive roots of all primes between 1 and 100,000.}
\label{fig:primitive_roots_all}
\end{figure}

Figure~\ref{fig:primitive_roots_smallest} graphs the smallest
primitive roots of all primes between 1 and 100000. The $x$-axis
represents primes between 1 and 100000. The $y$-axis represents the
smallest primitive root of each prime within that
interval. Figure~\ref{fig:primitive_roots_largest} shows a
corresponding graph for the largest primitive root of each prime
within the above interval.

\begin{figure}[!htbp]
\centering
\includegraphics{figures/primitive-roots-smallest}
\caption{The smallest primitive roots of all primes between 1 and 100,000.}
\label{fig:primitive_roots_smallest}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics{figures/primitive-roots-largest}
\caption{The largest primitive roots of all primes between 1 and 100,000.}
\label{fig:primitive_roots_largest}
\end{figure}
